Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-24 demystify.tex
updates

\subsection{Generic Letters are Bad} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.

@@

\subsection{Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. However, you can help your writer to write a strong, customized LoR for you by providing details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\subsection{Kiss of Death}

\begin{itemize}
\item \textbf{Too personal} Don't talk about your personal issues, e.g., family, health, relationship, etc. This can raise concerns about your ability to handle the PhD. Don't talk about religious or political beliefs (\autoref{sec:cultural}).
\item \textbf{Criticizing} your current or previous institution or professors: just like in a job interview, don't badmouth your current or previous employer. 
\item \textbf{No concrete evidence to back up claims}: e.g., saying you are passionate about a research topic without showing any experience. This is where specific names (work submitted at conf. X), numbers (outperformed SOTA by Y \%), and examples (worked on project Z) can help. 
\item \textbf{Use flowery and AI-like language}: don't use AI to write your SOP. Not that hard to raise suspicion.
\item \textbf{Not customized to the program}: If your SOP can be sent to multiple programs with few changes, it is too generic. Do some homework and mention why you want to spend the next 7 years there.
\item \textbf{Mentioning wrong professors}: do not mention emeritus professors or those who have left.  Do your homework to see if the professors you mention are still active in research.  Teaching and adjunct faculty are often not active in mentoring PhD students (\autoref{sec:faculty-types}).
\item \textbf{Too Long and fancy format}: As mentioned, keep it under 2 pages. Don't use too much coloring or fancy fonts (like those in Words). CS academics like using LaTeX (common way to write our papers and other documents), so write your SOP using LaTeX. 
\end{itemize}

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at Mason, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

\subsection{Grades} Grades generally do not matter much for CS PhD admission. PhD in CS is a research degree and doing well in undergrad courses does not necessarily mean you can do research (other factors such as research experiences, pubs, LoRs are more important).

@@

As with school reputation, you and your LoR writers can mention the grading system of your university if you think that is helpful for adcom to evaluate.

@@

\begin{commentbox}[Vu:]
  The minimum for Mason (being above this might not mean much, but below is a \red{red flag}).
  \begin{itemize}
    \item GPA: $\ge 3.0$ in your undergrad (but we also consider the rank/prestige of your school)
    \item GRE: not required 
    % \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
    \item English requirement tests (one of the below)
          \begin{itemize}
            \item TOEFL: 88 pts in total AND $\ge 20$ points in each subsection OR
            \item IELTS: $\ge 6.5$ OR
            \item DuoLingo Graduate English: $\ge 120$ OR
            \item Pearson Test of Academic English: $\ge 67$
          \end{itemize}
  \end{itemize}
\end{commentbox}

\subsection{Online Courses and Certificates}
In general, these do not carry much weight as they do not show research ability. We do not care much if you have taken an online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you might be able to use these to show you have sufficient CS knowledge.

@@

By around March you should hear back from most PhD programs that want to admit you.  In many cases you will not hear back from schools that do not want to admit you (see \autoref{sec:late-rejection}).
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

Many students lament that they get no interviews or are rejected (with strong profiles) and that the admission process seems random.  However, while it is true that the process is not perfect, it is not random.
Other than not having a strong profile (e.g., research potential, GPA, LoRs, SOP, poor interviews), here are some other reasons for rejections.

@@

A related reason is that you did not show enough interest in the program.
For example, your SOP did not mention any faculty or research area at the program, you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges), or during the interview you showed little knowledge, interest, or enthusiasm (because they are your safety schools) in the program or profs.

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. In your application, indicate that you need financial assistance. Typically, adcom will either admit you and give you TA, or reject you. We do not admit a student without supporting them.

@@

Even if you have other funding and do not need a TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for a research career where you often give talks and tell people about your work. Mason sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and also get paid more (as a lecturer).

@@

Fellowship is another type of funding that students can get (e.g., from the university, industries, or government).
Fellowships are typically competitive and generous, giving pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipends (including summer) and open doors for job opportunities such as internships.

@@

\begin{center}
  \includegraphics[scale=0.3]{files/c6.png}
\end{center}

@@

\subsection{Finding the right adviser}

Here is some advice for finding a prof.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research about that prof. (e.g., go to their website) and contact them (\autoref{sec:contact}).

@@

\paragraph{Types of Faculty} There are typically two types of faculty at an R1 research university: tenure-line and teaching.  The main difference is their focus: research, teaching, or both. Tenure-line faculty are typically more research-focused and can mentor PhD students, while teaching faculty are more teaching-focused and typically do not advise PhD students. Thus when contacting a prof. for research opportunities (\autoref{sec:contact}), you should focus on tenure-line faculty (and not adjunct or emeritus :p).

@@

\paragraph{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. Typically, a tenure-track faculty starts with being an assistant professor and then gets tenure and promoted to associate (around 6 years) and then full (time varies, some become full within 3--4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with Assistant, then is promoted to Associated, and finally Full.

@@

Now that you know a bit about tenured and tenure-track faculty (\autoref{sec:faculty-types}).  Which one should you choose as your adviser? The short answer is that either can be a good adviser, but they have different strengths and weaknesses.

Tenure-track faculty, e.g., assistant professors, are likely young and active in research (they have to, to get tenure). Thus, they will devote more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

@@

\subsection{Things to avoid when cold emailing a professor}

\begin{itemize}

    \item \textbf{Not customizing the email}. These emails typically talk about the student and not why they are interested in the professor's work. See writing tips in \autoref{sec:contact}
    \item \textbf{Sending the same email to multiple professors}. Sending emails to multiple profs. using BCC or even CC (where we can see all addresses the email is being sent to) will result in no reply or a very harsh one on how unprofessional this is.
    \item \textbf{Too much information}. Keep your email short and to the point. If we are interested, we will ask for more information. 2--3 paragraphs are enough.  Personally I also do not want to see course transcript or CV in the first email.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email.  Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Not reading the professor's website} and asking questions that are already answered there.
    \item \textbf{Not using plain text}. Fancy format, colors, or fonts are not professional. CS people like plain text.
\end{itemize}

>>>>

\subsection{Generic Letters are Bad}\label{sec:generic-letters} 
When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.

@@

\subsection{Help your LOR Writer}]\label{sec:help-your-LOR-writers}

As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\subsection{Kiss of Death in SOP} \label{sec:kiss-of-death-sop}

\begin{itemize}
\item \textbf{Too personal} Don't talk about your personal issues, e.g., family, health, relationship, etc. This can raise concerns about your ability to handle the PhD. Don't talk about religious or political beliefs (\autoref{sec:cultural}).
\item \textbf{Criticizing} your current or previous institution or professors: just like in a job interview, don't badmouth your current or previous employer. 
\item \textbf{No concrete evidence to back up claims}: e.g., saying you are passionate about a research topic without showing any experience. This is where specific names (work submitted at conf. X), numbers (outperformed SOTA by Y \%), and examples (worked on project Z) can help. 
\item \textbf{Use flowery and AI-like language}: don't use AI to write your SOP. Not that hard to raise suspicion.
\item \textbf{Not customized to the program}: If your SOP can be sent to multiple programs with few changes, it is too generic. Do some homework and mention why you want to spend the next 5--7 years there.
\item \textbf{Mentioning wrong professors}: do not mention emeritus professors or those who have left.  Do your homework to see if the professors you mention are still active in research.  Teaching and adjunct faculty are often not active in mentoring PhD students (\autoref{sec:faculty-types}).
\item \textbf{Too Long and fancy format}: As mentioned, keep it under 2 pages. Don't use too much coloring or fancy fonts (like those in Words). CS academics like using LaTeX (common way to write our papers and other documents), so write your SOP using LaTeX. 
\end{itemize}

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at Mason, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}
\subsection{Grades}\label{sec:gpa}

Grades generally do not matter much for CS PhD admission. PhD in CS is a research degree and doing well in undergrad courses does not necessarily mean you can do research (other factors such as research experiences, pubs, LoRs are more important).

@@

As with school reputation, you and your LoR writers can mention the grading system of your university if you think that is helpful for adcom to evaluate (\autoref{sec:help-your-LOR-writers}).

@@

\begin{commentbox}[Vu:]
  Here is the minimum requirements at Mason. 
  Being above this might not mean much, but below is a \red{red flag}.
  \begin{itemize}
    \item GPA: $\ge 3.0$ in your undergrad (but we also consider the rank/prestige of your school)
    \item GRE: not required 
    % \item but if you want to use it, then we expect a total (V+Q) of $\ge 311$ (with a $\ge 157$ Q) and A $\ge 3.0-3.5$.
    \item \href{https://www.gmu.edu/international/english-language-requirements}{English proficiency requirements} (one of the below)
          \begin{itemize}
            \item TOEFL: 80 OR
            \item IELTS: $\ge 6.5$ OR
            \item DuoLingo Graduate English: $\ge 120$ OR
            \item Pearson Test of Academic English: $\ge 67$
          \end{itemize}
  \end{itemize}
\end{commentbox}

\subsection{Online Courses and Certificates}

These do not carry much weight as they do not show research ability. We do not care much if you have taken an online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you might be able to use these to show you have sufficient CS knowledge.

@@

By around March you should hear back from most PhD programs that want to admit you. 
But you likely won't hear back from schools that do not want to admit you (\autoref{sec:late-rejection}).

If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

@@

Many students lament that they get no interviews or are rejected and that the admission process seems random.  However, while it is true that the process is not perfect, it is not random.
Other than not having a strong profile (e.g., research potential, GPA, LoRs, SOP, poor interviews), here are some other reasons for rejections.

@@

A related reason is that you did not show enough interest in the program.
For example, you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges), or during the interview you showed little knowledge, interest, or enthusiasm in the profs or program (because it is your safety).

@@

\paragraph{How to get TA?}  Unless you have other funding such as RA or fellowships, TA is typically the default for CS PhD programs. In your application, just simply indicate that you need financial assistance. Typically, adcom will either admit you and give you TA, or reject you. We do not admit a student without supporting them (\autoref{sec:ievaluate}).

@@

Even if you have other funding and do not need a TA, you still should do TA at least once.  This allows you to see what teaching is like, which is especially helpful for a research career where you often give talks and tell people about your work. Mason sometimes has classes that a more senior student can teach.  In that case, you will be paid as GTA or even sometimes as a lecturer.  This is a good opportunity for students to get teaching experience and you might even get paid a bit more.

@@

Fellowship is another type of funding that students can get from the university, industries, or government.
Fellowships are typically competitive and generous, giving pretty much all benefits tuition/insurance that a TA/RA has.  They might even give higher stipends (including summer) and open doors for job opportunities such as internships.

@@

\begin{center}
  \includegraphics[scale=0.5]{files/c6.png}
\end{center}

@@

\subsection{Finding an Adviser}

Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research about that prof. (e.g., go to their website) and contact them (\autoref{sec:contact}).

@@

\subsubsection{Types of Faculty} There are typically two types of faculty at an R1 research university: tenure-line and teaching.  The main difference is their focus: research, teaching, or both. Tenure-line faculty are typically more research-focused and can mentor PhD students, while teaching faculty are more teaching-focused and typically do not advise PhD students. Thus when contacting a prof. for research opportunities (\autoref{sec:contact}), you should focus on tenure-line faculty (and not adjunct or emeritus :p).

@@

\subsubsection{Ranks} Faculty have ranks such as assistant, associate, and full, regardless if they are tenure-line or teaching.  Assistant means new faculty, associate means they have been promoted (or was hired after being in the industry for a while), and full means they are senior.

Typically, a tenure-track faculty starts with being an assistant professor and then gets tenure and promoted to associate (around 6 years). Note that getting tenure is a huge deal, and it might change the way they work with students, e.g., they might be more relaxed and less stressed (\autoref{sec:tenure-vs-tenure-track}).

After that they go up for full prof. (time varies, some become full within 3--4 years, some 10+ years, some remain associate). A non-tenure track faculty, e.g., teaching or research faculty, also starts with Assistant, then is promoted to Associated, and finally Full.

@@

Now that you know a bit about tenured and tenure-track faculty (\autoref{sec:faculty-types}).  Which one should you choose as your adviser? 
Either can be a good adviser, but they have different strengths and weaknesses.

Tenure-track faculty, e.g., assistant professors, are likely young and active in research (they have to, to get tenure). Thus, they will devote more time to work with you and push you to do research and publish. However, you might not be too independent when you graduate because they have been too hands-on with you.  Also, they may not have as much experience in managing students and may not have as much funding (yet).

@@

\subsection{Kiss of Death in Emails}\label{sec:kiss-of-death-emails}

\begin{itemize}
    \item \textbf{Not customizing the email, too generic}: a copy and paste kind of email or those that can be sent to multiple professors with very little modifications. This shows the lack of interest and will be treated as spam.  
    \item \textbf{Self-focus} You focus too much about you and your achievements but not why you are interested in the prof.'s work (\autoref{sec:contact}). Why their work is interesting and how your background fits in.
    \item \textbf{Too long}. Keep it to about 3--4 paragraphs. Seriously, less is more and long emails are often not read and discarded.\\
    Personally I do not want to see course transcript or CV in the first email.  If we are interested, we will ask for more information.
    \item \textbf{Flowery greetings and language}. Don't use ``Dear esteemed professor'', ``kindly consider my application'', etc. And of course, don't ``hey there''. 
    \item \textbf{Sending the similar mail to multiple profs}. Sending emails to multiple profs. using BCC or even CC (where we can see all addresses the email is being sent to) will result in no reply or a very harsh one on how unprofessional this is.
    \item \textbf{Not being professional}, e.g., do not call the professor by their first name in the first email (some don't care but you don't want to take the risk -- you don't know them that well yet).  Do not use Mr., Mrs., etc. Use Prof. Lastname or Dr. Lastname (\autoref{sec:address}).
    \item \textbf{Ignoring the Prof's guideline} and asking questions that are already answered on their website (most CS profs have personal websites). For example, on my website, I put very specific information on how to contact me (e.g., email subject).
    \item \textbf{Fancy format}. CS people like plain text.  Do not use colors, fonts, or fancy formats.
\end{itemize}
